---
title: "native_bee_data_analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!--
STRUCTURE OF THIS DOC:
  1. Setup
  2. Reference table (SD County taxonomic source of truth)
  3. Clean data (sources bee_specimen_clean.R + bee_inat_clean.R)
  4. Spatial (transects + buffers, via spatial_utils.R)
  5. ANALYSIS: taxon richness by geographic scale
  6. ANALYSIS: method comparison (lethal vs non-lethal)
  7. Plants (placeholder, TODO)

This doc is for ANALYSIS only. All cleaning lives in dedicated
scripts (scripts/bee_specimen_clean.R, scripts/bee_inat_clean.R)
so this stays focused on comparison, not data wrangling.
-->

#Questions
####Reminder for comparison surveys, you can only compare months through April/March-September because interns were not surveying throughout the entire year

#1. What bees exist in lethal vs. nonlethal (where do they not overlap, show a Venn Diagram)?
####1. a. Where do we need more collective sampling if we find bees in our nonlethal dataset that aren't in our lethal, and vice versa? For example, the most common Lasioglossum Dialictus are missing from the nonlethal set because pictures are not enough to get ID's on such small animals (use CABR checklists for this).

#2. What is the taxonomic resolution (taxon rank) you can get to for lethal vs. nonlethal? eg.) Nonlethal most commmonly only identifies to genus, and nonlethal most commonly gets species

#3. What top 10 plants to bees visit the most throughout the year and then per month? Compare by park and by nonlethal surveys. We want to know what plants we should plant more of.

#4. Network analysis what flowers to what bees visit to get those unique bees (rarer bee, what plant does it visit?).
####4. a. Need species rank for finding specialist relationships 

#5. Have we accumulated all the species there is to find at Cabrillo? (Graph a species accumulation curve)

#6.Are there bees in the park that weren't found on the transects? Do we need more transects? (suveyors vs. whole park) Which bees were those? What are the observations url's associated with them to see what plants they were on?

#7. What needs ID in our lethal and non-lethal specimens most frequently? (Target identification)

#8. Across the years, how has diversity changed, in genus-->species, and how has it changed monthby month? Species diversity versus genus/subgenus diversity? Where is the most diversity concentrated in the park for bees?

#10. What bees are in CABR Official Checklist that isn't in the Holway Official Checklist?

#11. What surveys captured more bees by species rank, beeple vs. interns, lethal vs. nonlethal?

#12. Give plant phenology based on the survey data (nonlethal only)













###### 1. SETUP ######
```{r setup, message=FALSE, warning=FALSE}
# Run these ONCE to install, then leave commented out.
# install.packages(c("tidyverse", "dplyr", "ggplot2", "writexl", "egg",
#                    "stringr", "lubridate", "rgbif", "gbifdb", "sf", "httr2", "readxl"))
# remotes::install_github("duckdb/duckdb")

library(tidyverse)   # data cleaning + ggplot2, dplyr, stringr, etc.
library(writexl)     # write Excel files
library(readxl)      # read Excel files
library(egg)         # arranging ggplot figures
library(lubridate)   # parse dates, pull month/day/year
library(rgbif)       # GBIF access (optional, for later)
library(gbifdb)      # GBIF remote database (optional, for later)
library(sf)          # spatial data (shapefiles, transects)
library(httr2)       # iNaturalist API calls (used by sourced scripts)

source("scripts/utils/utils.R")  # read_latest(), require_columns()
```

###### 2. TIERED CHECKLISTS (build if missing — calls iNat API, ~3-4 min) ######
```{r build-checklists, message=FALSE}
# native_bee_checklist.R makes ~430 iNaturalist API calls (~3-4 min),
# so we only re-run it if the saved checklist CSVs don't already exist.
# Delete any of the three data/checklists/*_inat_native_bee_checklist.csv
# files to force a refresh of all three (they're built together,
# from one shared API pass over the union of taxa).

if (!file.exists("data/checklists/sd_county/sd_county_inat_native_bee_checklist.csv") ||
    !file.exists("data/checklists/point_loma/pl_inat_native_bee_checklist.csv") ||
    # RENAMED 2026-06-24 (was CABR_inat_native_bee_checklist.csv)
    !file.exists("data/checklists/cabr/cabr_inat_bee_checklist.csv")) {
  message("One or more tiered checklists not found - running native_bee_checklist.R (this calls the API, ~3-4 min)...")
  source("scripts/checklists/native_bee_checklist.R")
}

```

###### 3. CLEAN DATA ######
```{r clean-data, message=FALSE}
# Produces clean_bee_data (lethal specimens).
source("scripts/specimens/specimen_bee_clean.R")

# Produces clean_intern_data, clean_beeple_data, and the combined
# clean_nonlethal_data. Will report "no data found yet" until intern/
# beeple iNat exports exist — that's expected, not an error.
source("scripts/observations/inat_bee_clean.R")
```

###### 4. SPATIAL (transects + buffers) ######
```{r spatial}
# Produces `transects` and `buffer_50m` in the environment.
source("scripts/spatial/spatial_utils.R")

trails <- c("BST", "UPMON", "TP", "OT")
```

###### 5. ANALYSIS: TAXON RICHNESS BY GEOGRAPHIC SCALE ######
```{r richness-by-scale}
# The 6 bee families found in San Diego County (Stenotritidae is
# Australia-only and is not expected here - kept for reference).
bee_family <- c("Andrenidae", "Apidae", "Colletidae",
                "Halictidae", "Megachilidae", "Melittidae")

# bee_reference_sd_county / _point_loma / _cabr each have key_full,
# key_genus_species, key_species_only, rank, and full taxonomy --
# built in section 2, one per geographic tier.
sd_taxon_richness         <- n_distinct(bee_reference_sd_county$key_full)
point_loma_taxon_richness <- n_distinct(bee_reference_point_loma$key_full)
cabr_taxon_richness       <- n_distinct(bee_reference_cabr$key_full)

cat("San Diego County taxon richness:", sd_taxon_richness, "\n")
cat("Point Loma taxon richness:      ", point_loma_taxon_richness,
    sprintf(" (%.1f%% of SD County)\n", 100 * point_loma_taxon_richness / sd_taxon_richness))
cat("CABR taxon richness:            ", cabr_taxon_richness,
    sprintf(" (%.1f%% of SD County)\n", 100 * cabr_taxon_richness / sd_taxon_richness))

# NOTE: these are TIER 1 (iNat-only) richness figures. They will
# change once TIER 2 (merged with specimen + Dorey data) checklists
# exist -- see README "checklist architecture" TODO.
```

###### 6. ANALYSIS: METHOD COMPARISON (LETHAL VS NON-LETHAL) ######
```{r method-comparison}
lethal_richness <- n_distinct(clean_bee_data$oldscientificname)
cat("Lethal (specimen) richness at CABR:", lethal_richness, "\n")

if (exists("clean_nonlethal_data") && nrow(clean_nonlethal_data) > 0) {
  nonlethal_richness <- n_distinct(clean_nonlethal_data$taxon_id)
  cat("Non-lethal (photo) richness at CABR:", nonlethal_richness, "\n")
} else {
  cat("Non-lethal richness: pending intern/beeple data.\n")
}

# TODO: once non-lethal data exists, compare overlap using taxon_id
# (exact match) first, falling back to complex for
# complex-level photo IDs that can't be resolved to species
# (see native_bee_checklist.R).
```

###### 7. PLANTS (placeholder) ######
```{r plant-checklist, eval=FALSE}
# TODO: build once plant checklist script (TODO #5) exists.
# Column naming note: follows the same bare-name convention as the
# bee pipeline (2026-06-24 decision) -- family/genus, not
# taxon_family_name/taxon_genus_name. taxon_id and taxon_rank are
# untouched -- different fields, not part of that rename.
# plants_path <- read_latest("data/cabr_surveys/nonlethal/inat_plant", "^inat_plants_point_loma_peninsula")
# sd_county_plant_checklist <- read_csv(plants_path) %>%
#   filter(taxon_rank %in% c("genus", "subgenus", "species", "subspecies")) %>%
#   distinct(taxon_id, .keep_all = TRUE) %>%
#   select(taxon_id, taxon_rank, scientific_name,
#          family, genus) %>%
#   arrange(family, genus, taxon_rank, scientific_name)
```
